---
title: "ST_558_HW8"
author: "Jay Thakur"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---
# ST 558 HW8

## Reading Data

##### Loading all req libraries
```{r}
library(tidymodels)
library(dplyr)
library(ggplot2)
library(lubridate)
library(readr)
```

#### Reading through the data
Using read.csvto read data and as I was having issue with header I decided to drop and reassign names to headers (Same dataset was used for my Project 2)

```{r}
# Read the CSV file
df <- read.csv("SeoulBikeData.csv", header = FALSE, stringsAsFactors = FALSE)

# Remove the first row
df <- df[-1, ]
colnames(df) <- c("date", "rented_bike_count", "hour", "temperature", "humidity", 
                  "wind_speed", "visibility", "dew_point_temp", "solar_radiation",
                  "rainfall", "snowfall", "seasons", "holiday", "functioning_day")

```

## EDA

#### Checking for missing values if any

```{r}
missing_summary <- df |>
  summarise(across(everything(), ~ sum(is.na(.))))
print(missing_summary)
```

#### Check column types and summary
```{r}
glimpse(df)
```

#### Summary statistics for numeric and categorical columns
```{r}
numeric_summary <- df |>
  summarise(across(where(is.numeric), list(min = min, mean = mean, max = max), na.rm = TRUE))
print(numeric_summary)
```

```{r}
categorical_summary <- df |>
  summarise(across(where(is.character), ~ list(unique(.))))
print(categorical_summary)
```

#### Converting columns to appropriate type
Converting date to Date type and Season, Holiday and Functioning day to factor. Rest of the columns should be numeric.
```{r}
df <- df |>
  mutate(
    date = as.Date(date, format = "%d/%m/%Y"),
    seasons = factor(seasons),
    holiday = factor(holiday),
    functioning_day = factor(functioning_day)
  )

df <- df |>
  mutate(across(c(rented_bike_count, hour, temperature, humidity, wind_speed,
                  visibility, dew_point_temp, solar_radiation, 
                  rainfall, snowfall), as.numeric))

```

#### Summary statistics for numeric and categorical columns
Summarizing after type conversion.

```{r}
numeric_summary <- df |>
  summarise(across(where(is.numeric), list(min = min, mean = mean, max = max), na.rm = TRUE))
print(numeric_summary)
```

```{r}
categorical_summary <- df |>
  summarise(across(where(is.factor), ~ list(levels(.))))
print(categorical_summary)
```

#### Generating statistic for Bike count
```{r}
functioning_data <- df

functioning_summary <- functioning_data |>
  group_by(seasons, holiday, functioning_day) |>
  summarise(
    mean_bike_count = mean(rented_bike_count),
    total_bike_count = sum(rented_bike_count),
    .groups = "drop"
  )
print(functioning_summary)

```

From above we can see that Bike renting was Zero when Functioning day is "No". Hence we can drop those rows which are useless.

```{r}
functioning_data <- df |>
  filter(functioning_day == 'Yes')

functioning_summary <- functioning_data |>
  group_by(seasons, holiday, functioning_day) |>
  summarise(
    mean_bike_count = mean(rented_bike_count),
    total_bike_count = sum(rented_bike_count),
    .groups = "drop"
  )
print(functioning_summary)
```

#### Summarizing Data Across Hours to daily observations
Summarizing by grouping with seasons and Holiday across hours.

```{r}
df_daily <- functioning_data |>
  group_by(date, seasons, holiday) |>
  summarise(
    total_bike_count = sum(rented_bike_count),
    total_rainfall = sum(rainfall),
    total_snowfall = sum(snowfall),
    mean_temperature = mean(temperature),
    mean_humidity = mean(humidity),
    mean_wind_speed = mean(wind_speed),
    mean_visibility = mean(visibility),
    mean_dew_point_temp = mean(dew_point_temp),
    mean_solar_radiation = mean(solar_radiation),
    .groups = "drop"
  )
```

```{r}
summary(df_daily)
```

```{r}
# Verifying that 'date' is of class Date
class(df_daily$date)
```

#### Plotting Renting bike count across temperature

```{r}
ggplot(df_daily, aes(x = mean_temperature, y = total_bike_count)) +
  geom_point() +
  labs(x = "Mean Temperature (°C)", y = "Total Bike Count", title = "Bike Rentals vs. Temperature")

```

#### Displaying correlation matric across variables

```{r}
numeric_vars <- df_daily |>
  select(total_bike_count, total_rainfall, total_snowfall, mean_temperature, mean_humidity, mean_wind_speed, mean_visibility, mean_dew_point_temp, mean_solar_radiation)

cor_matrix <- cor(numeric_vars, use = "complete.obs")
print(cor_matrix)

```

## Split the Data

#### Splitting into Training and Test Sets
Using 75/25 split to split data into Training and Testing.
```{r}
set.seed(123)
data_split <- initial_split(df, prop = 0.75, strata = seasons)
train_data <- training(data_split)
test_data <- testing(data_split)
```

#### Creating 10 cross-validation folds for training data which will be used for fitting and validation.

```{r}
cv_folds <- vfold_cv(train_data, v = 10)
```

## Fitting MLR Models

#### Creating Recipe 1
Creating day of week variable to get weekday and weekend, converting cat varibles to dummy and normalizing numric variables.

```{r}

recipe1 <- recipe(rented_bike_count ~ ., data = train_data) |>
  step_date(date, features = c("dow"), label = FALSE) |>
  step_mutate(
    day_type = factor(
      if_else(date_dow %in% c(1, 7), "weekend", "weekday"),
      levels = c("weekday", "weekend")
    )
  ) |>
  step_rm(date, date_dow) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors()) |>
  step_lincomb(all_predictors())

prepped_recipe <- prep(recipe1, training = train_data)
baked_data <- bake(prepped_recipe, new_data = NULL)
glimpse(baked_data)

```

#### Creating Recipe 2
Adding interatction between mentioned variables.

```{r}
recipe2 <- recipe(rented_bike_count ~ ., data = train_data) |>
  step_date(date, features = c("dow"), label = FALSE) |>
  step_mutate(
    day_type = factor(
      if_else(date_dow %in% c(1, 7), "weekend", "weekday"),
      levels = c("weekday", "weekend")
    )
  ) |>
  step_rm(date, date_dow) |>
  step_interact(terms = ~ starts_with("seasons"):starts_with("holiday")) |>
  step_interact(terms = ~ temperature:starts_with("seasons")) |>
  step_interact(terms = ~ temperature:rainfall) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors()) |>
  step_lincomb(all_predictors())

prepped_recipe <- prep(recipe2, training = train_data)
baked_data <- bake(prepped_recipe, new_data = NULL)
glimpse(baked_data)
```

#### Creating Recipe 3
Adding 2nd degree poly term for each numric variable.

```{r}
numeric_vars <- c(
  "hour",
  "temperature",
  "humidity",
  "wind_speed",
  "visibility",
  "dew_point_temp",
  "solar_radiation",
  "rainfall",
  "snowfall"
)

recipe3 <- recipe2 |>
  step_poly(all_of(numeric_vars), degree = 2)

prepped_recipe <- prep(recipe3, training = train_data)
baked_data <- bake(prepped_recipe, new_data = NULL)
glimpse(baked_data)
```

#### Setting up Linear regression model

```{r}
lm_model <- linear_reg() |>
  set_engine("lm")
```


```{r}
workflow1 <- workflow() |>
  add_model(lm_model) |>
  add_recipe(recipe1)

workflow2 <- workflow() |>
  add_model(lm_model) |>
  add_recipe(recipe2)

workflow3 <- workflow() |>
  add_model(lm_model) |>
  add_recipe(recipe3)
```

#### Fitting training data on each model of receipe

```{r}
set.seed(123)
results1 <- fit_resamples(
  workflow1,
  resamples = cv_folds,
  metrics = metric_set(rmse)
)

results2 <- fit_resamples(
  workflow2,
  resamples = cv_folds,
  metrics = metric_set(rmse)
)

results3 <- fit_resamples(
  workflow3,
  resamples = cv_folds,
  metrics = metric_set(rmse)
)
```

#### Analyzing performance metric of each model

```{r}
collect_metrics(results1)
collect_metrics(results2)
collect_metrics(results3)
```

#### Choice of Final Model

As per above metric we can see that Third model has less RMSE value and standard error value compared to other two, Hence I will be using third model for fitting whole training set and testing.

```{r}
final_fit <- last_fit(workflow3, split = data_split)
```

```{r}
final_fit |>
  collect_metrics()
```

As we can see RMSE value we get here is less than CV RMSE of all 3 models. Though R-square value indicates mid-level generalization of data.

```{r}
final_model <- extract_fit_parsnip(final_fit$.workflow[[1]])

tidy(final_model)
```



